---
output:
  xaringan::moon_reader:
    css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

layout: true

.hheader[<a href="index.html">`r fontawesome::fa("home", fill = "steelblue")`</a>]

---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
knitr::opts_chunk$set(fig.height=5, fig.align="center")
library(huxtable)
```

class: center, middle, inverse
# Seasonal ARIMA Models

.futnote[Eli Holmes, UW SAFS]

.citation[eeholmes@uw.edu]

---

```{r load_data, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)
library(reshape2)
library(tseries)
library(forecast)
```

## Seasonality

Load the chinook salmon data set

```{r}
load("chinook.RData")
head(chinook)
```

---

The data are monthly and start in January 1990.  To make this into a ts object do

```{r}
chinookts <- ts(chinook$log.metric.tons, start=c(1990,1), 
                frequency=12)
```
`start` is the year and month and frequency is the number of months in the year.  

Use `?ts` to see more examples of how to set up ts objects.

---

## Plot seasonal data

```{r}
plot(chinookts)
```

---

## Seasonal ARIMA model

Seasonally differenced data:

$$z_t = x_t - x_{t+s} - m$$

Basic structure of a seasonal AR model

$z_t$ = AR(p) + AR(season) + AR(p+season)


Example AR(1) non-seasonal part + AR(1) seasonal part

$$z_t = \phi_1 z_{t-1} + \Phi_1 z_{t-12} - \phi_1\Phi_1 z_{t-13}$$

---

## Notation

ARIMA (p,d,q)(ps,ds,qs)S


ARIMA (1,0,0)(1,1,0)[12]

Notice we are modeling $x$ this year in Jan (say) as a function of $x$ in Jan last year.

---

## Seasonal models

Let's imagine that we can describe our data as a combination of the mean trend, a seasonal term, and error.

$$x_t = \mu t+ s_t + w_t$$
Let's imagine that the seasonal term is just a constant based on month and doesn't change with time.

$$s_t = f(month)$$

---

We want to remove the $s_t$ with differencing so that we can model $e_t$.  We can solve for $x_{t+1}$ by using $x_{t-s}$ where $s$ is the seasonal length (e.g. 12 if season is yearly).

When we take the first seasonal difference, we get
$$\Delta_s x_t = \mu(t-(t-s)) + s_t - s_{t-s} + w_t - w_{t-s} = \mu s + w_t - w_{t-s}$$
The $s_t-s_{t-s}$ disappears because $s_t = s_{t-s}$ when the seasonal effect is just a function of the month.  Depending on what $m_t$ is, we might be done or we might have to do a first difference.  Notice that the error term is a moving average in the seasonal part.

---

```{r, echo=FALSE}
set.seed(123)
yr=10
s=12
sd=sqrt(.1)
mu=0.05
st=sin(pi*(1:(s*yr))/(s/2))
mt=mu*(1:(s*yr))
et=rnorm(s*yr,0,sd)
xt=mt+st+et
xt=ts(xt,start=1,frequency=12)
plot(xt)
```

```{r}
plot(diff(xt,lag=12))
```

---

We can recover the model parameters with `Arima()`. Note the drift term is returned at $\mu$ not $\mu s$.
```{r}
forecast::Arima(xt, seasonal=c(0,1,1), include.drift=TRUE)
```

---

`auto.arima()` identifies a ARIMA(0,0,0)(0,1,2)[12].
```{r}
forecast::auto.arima(xt, stepwise=FALSE)
```

---

## Seasonal model with changing season

Let's imagine that our seasonality is increasing over time.

$$s_t = \beta \times year \times f(month)\times$$
When we take the first seasonal difference, we get
$$\Delta_s x_t = \mu(t-(t-s)) + \beta f(month)\times (year - (year-1)) + w_t - w_{t-s} \\ = \mu s + \beta f(month) + w_t - w_{t-s}$$

---

We need to take another seasonal difference to get rid of the $f(month)$ which is not a constant; it is different for different months as it is our seasonality.
$$\Delta^2_{s} x_t = w_t - w_{t-s} - w_{t-s} + w_{t-2s}=w_t - 2w_{t-s} + w_{t-2s}$$
So our ARIMA model should be ARIMA(0,0,0)(0,2,2)

---

```{r, echo=FALSE}
set.seed(123)
yr=10
s=12
sd=sqrt(.1)
mu=0.05
beta=.2
st=(1+beta*rep(1:yr, each=s))*sin(pi*(1:(s*yr))/(s/2))
st=beta*rep(1:yr, each=s)*sin(pi*(1:(s*yr))/(s/2))
mt=mu*(1:(s*yr))
et=rnorm(s*yr,0,sd)
xt=mt+st+et
xt=ts(xt,start=1,frequency=12)
plot(xt)
```

---

But we can recover the model with `Arima()`. Note the drift term is returned at $\mu$ not $\mu s$.
```{r}
forecast::Arima(xt, seasonal=c(0,2,2))
```

---

`auto.arima()` again has problems and returns many Infs; turn on `trace=TRUE` to see the problem.
```{r}
forecast::auto.arima(xt, stepwise=FALSE)
```

---

## Seasonal model with changing season #2

Let's imagine that our seasonality increases and then decreases.

$$s_t = (a y^2-b y+h) f(month)$$
```{r, echo=FALSE}
set.seed(100)
yr=10
s=12
sd=sqrt(.1)
mu=0.05
beta=.1*(rep(1:yr, each=s)^2-2*5*rep(1:yr, each=s)+30)
st=beta*sin(pi*(1:(s*yr))/(s/2))
mt=mu*(1:(s*yr))
et=rnorm(s*yr,0,sd)
xt=mt+st+et
xt=ts(xt,start=1,frequency=12)
plot(xt)
```

---

Then we need to take 3 seasonal differences to get rid of the seasonality. The first will get rid of the $h f(month)$, the next will get rid of $by$ (year) terms and $y^2$ terms, the third will get rid of extra $y$ terms introduced by the 2nd difference.  The seasonal differences will get rid of the linear trend also.

$$\Delta^3_{s} x_t = w_t - 2w_{t-s} + w_{t-2s}-w_{t-s}+2w_{t-2s}-w_{t-3s}=w_t - 3w_{t-s} + 3w_{t-2s}-w_{t-3s}$$

So our ARIMA model should be ARIMA(0,0,0)(0,3,3).

---

## `auto.arima()` for seasonal ts

`auto.arima()` will recognize that our data has season and fit a seasonal ARIMA model to our data by default.  We will define the training data up to 1998 and use 1999 as the test data.

```{r}
traindat <- window(chinookts, c(1990,10), c(1998,12))
testdat <- window(chinookts, c(1999,1), c(1999,12))
fit <- forecast::auto.arima(traindat)
fit
```

---

## Forecast using seasonal model

```{r}
fr <- forecast::forecast(fit, h=12)
plot(fr)
points(testdat)
```

---

## Missing values

Missing values are ok when fitting a seasonal ARIMA model

```{r echo=FALSE}
plot(fr)
```

---

## Summary

Basic steps for identifying a seasonal model.  **forecast** automates most of this.

* Check that you have specified your season correctly in your ts object.

* Plot your data.  Look for trend, seasonality and random walks.

---

## Summary

* Use differencing to remove season and trend.

    * Season and no trend.  Take a difference of lag = season
    * No seasonality but a trend.  Try a first difference
    * Both. Do both types of differences
    * Neither. No differencing.
    * Random walk. First difference.
    * Parametric looking curve. Tranform.
    
---

## Summary

* Examine the ACF and PACF of the differenced data.
    * Look for patterns (spikes) at seasonal lags

* Estimate likely models and compare with model selection criteria (or cross-validation). Use `TRACE=TRUE`

* Do residual checks. 

