---
output:
  xaringan::moon_reader:
    css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

layout: true

.hheader[<a href="index.html">`r fontawesome::fa("home", fill = "steelblue")`</a>]

---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
knitr::opts_chunk$set(fig.height=5, fig.align="center")
library(huxtable)
```

class: center, middle, inverse
# ARIMA Models
## Forecasting

.futnote[Eli Holmes, NOAA Fisheries]

.citation[eli.holmes@noaa.gov]

---

```{r load_data, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)
library(reshape2)
library(tseries)
library(forecast)
```

```{r load_data2, message=FALSE, warning=FALSE, echo=FALSE}
load("landings.RData")
landings$log.metric.tons = log(landings$metric.tons)
landings = subset(landings, Year <= 1987)
anchovy = subset(landings, Species=="Anchovy")$log.metric.tons
sardine = subset(landings, Species=="Sardine")$log.metric.tons
```

## Forecasting

The basic idea of forecasting with an ARIMA model is the same as forecasting with a time-varying regressiion model.

We estimate a model and the parameters for that model.  For example, let's say we want to forecast with ARIMA(2,1,0) model:
$$y_t = \beta_1 y_{t-1} + \beta_2 y_{t-2} + e_t$$
where $y_t$ is the first difference of our anchovy data.

---

Let's estimate the $\beta$'s:

```{r}
fit <- Arima(anchovy, order=c(2,1,0))
coef(fit)
```

So we will forecast with this model:

$$y_t = -0.3348 y_{t-1} - 0.1454 y_{t-2} + e_t$$

So to get our forecast for 1988, we do this

$$(y_{1988}-y_{1987}) = -0.3348 (y_{1987}-y_{1986}) - 0.1454 (y_{1986}-y_{1985})$$

$$y_{1988} = y_{1987}-0.3348 (y_{1987}-y_{1986}) - 0.1454 (y_{1986}-y_{1985})$$

---

$$y_{1988} = y_{1987}-0.3348 (y_{1987}-y_{1986}) - 0.1454 (y_{1986}-y_{1985})$$

Here is R code to do that:

```{r}
anchovy[24]+coef(fit)[1]*(anchovy[24]-anchovy[23])+
  coef(fit)[2]*(anchovy[23]-anchovy[22])
```

Thankfully, `forecast()` automates this calculation for us.

```{r}
forecast(fit, h=1)
```

---

## Forecasting

We can forecast from a fit in R using the `forecast()` function.  `h` is the number of time steps forward to forecast.  The upper and lower prediction intervals are shown.

```{r}
fit <- auto.arima(sardine, test="adf")
fr <- forecast(fit, h=5)
fr
```

---

We can plot our forecast with prediction intervals.  Here is the sardine forecast:

```{r}
plot(fr, xlab="Year")
```

---

# Repeat for anchovy

```{r}
fit <- auto.arima(anchovy)
fr <- forecast(fit, h=5)
plot(fr)
```

---

# What happens if I have missing values?

```{r}
anchovy.miss <- anchovy
anchovy.miss[10:14] <- NA
fit <- auto.arima(anchovy.miss)
fr <- forecast(fit, h=5)
plot(fr)
```

---

# Repeat for Chub Mackerel

```{r}
dat <- subset(landings, Species=="Chub.mackerel")$log.metric.tons
fit <- auto.arima(dat)
fr <- forecast(fit, h=5)
plot(fr, ylim=c(6,10))
```


---

## The "Naive" forecast

The "naive" forecast is simply the last value observed.  If we want to prediction landings in 2019, the naive forecast would be the landings in 2018.  This is a difficult forecast to beat!  It has the advantage of having no parameters.

In forecast, we can fit this model with the `naive()` function.  Note this is the same as the `rwf()` function.

---

```{r}
fit.naive <- naive(anchovy)
fr.naive <- forecast(fit.naive, h=5)
plot(fr.naive)
```

---

## The "Naive" forecast with drift

The "naive" forecast is equivalent to a random walk with no drift.  So this
$$x_t = x_{t-1} + e_t$$
As you saw with the anchovy fit, it doesn't allow an upward trend.  Let's make it a little more flexible by add `drift`.  This means we estimate one term, the trend.

$$x_t = \mu + x_{t-1} + e_t$$

---

```{r}
fit.rwf <- rwf(anchovy, drift=TRUE)
fr.rwf <- forecast(fit.rwf, h=5)
plot(fr.rwf)
```

---

## The "mean" forecast

The "mean" forecast is simply the mean of the data.  If we want to prediction landings in 2019, the mean forecast would be the average of all our data.  This is a poor forecast typically.  It uses no information about the most recent values.

In forecast, we can fit this model with the `Arima()` function and `order=c(0,0,0)`.  This will fit this model:
$$x_t = e_t$$
where $e_t \sim N(\mu, \sigma)$.

---

```{r}
fit.mean <- Arima(anchovy, order=c(0,0,0))
fr.mean <- forecast(fit.mean, h=5)
plot(fr.mean)
```

