---
output:
  xaringan::moon_reader:
    css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

layout: true

.hheader[<a href="index.html">`r fontawesome::fa("home", fill = "steelblue")`</a>]

---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
library(huxtable)
library(ggplot2)
library(forecast)
```

class: center, middle, inverse
# Forecasting Time Series
## Time-varying Regression: Forecasting

.futnote[Eli Holmes, UW SAFS]

.citation[eeholmes@uw.edu]

---

```{r load_data_TV_Regression_Forecasting, echo=FALSE}
load("landings.RData")
landings$log.metric.tons = log(landings$metric.tons)
landings = subset(landings, Year <= 1989)
landings$t = landings$Year-landings$Year[1]
anchovy = subset(landings, Species=="Anchovy" & Year <= 1987)
sardine = subset(landings, Species=="Sardine" & Year <= 1987)
```

Forecasting is easy in R once you have a fitted model.  

Let's say for the anchovy, we fit the model

$$C_t = \alpha + \beta t + e_t$$
where $t$ starts at 0 (so 1964 is $t=0$ ).  To predict, predict the catch in year t, we use

$$C_t = \alpha + \beta t + e_t$$
---

Model fit:

```{r tvreg.anchovy2}
model <- lm(log.metric.tons ~ t, data=anchovy)
coef(model)
```

For anchovy, the estimated $\alpha$ (Intercept) is `r coef(model)[1]` and $\beta$ is `r coef(model)[2]`.  We want to use these estimates to forecast 1988 ( $t=24$ ).

So the 1988 forecast is `r coef(model)[1]` + `r coef(model)[2]` $\times$ 24 :

```{r tvreg.forecast1}
coef(model)[1]+coef(model)[2]*24
```

log metric tons.

---

# The forecast package

The forecast package in R makes it easy to create forecasts with fitted models and to plot (some of) those forecasts.

For a TV Regression model, our `forecast()` call looks like

```{r TVregression.forecast2}
library(forecast)
fr <- forecast(model, newdata = data.frame(t=24:28))
```

---

The dark grey bands are the 80% prediction intervals and the light grey are the 95% prediction intervals.

```{r plot.TVreg.forecast, fig.align = "center"}
plot(fr)
```

---

Sardine forecasts from a higher order polynomial can similarly be made.  Let's fit a 4-th order polynomial.


$$C_t = \alpha + \beta_1 t + \beta_2 t^2 + \beta_3 t^3 + \beta_4 t^4 + e_t$$

To forecast with this model, we fit the model to estimate the $\beta$'s and then replace $t$ with $24$:

$$C_{1988} = \alpha + \beta_1 24 + \beta_2 24^2 + \beta_3 24^3 + \beta_4 24^4 + e_t$$

---

This is how to do that in R:

```{r tvreg.sardine2}
model <- lm(log.metric.tons ~ t + I(t^2) + I(t^3) + I(t^4), data=anchovy)
fr <- forecast(model, newdata = data.frame(t=24:28))
fr
```

---

Unfortunately, forecast does not recognize that there is only one predictor $t$ and we cannot use forecast's plot function.

If you do this in R, it throws an error.
```{r plot.TVreg.forecast.bad}
try(plot(fr))
```

```
Error in plotlmforecast(x, PI = PI, shaded = shaded, shadecols = shadecols, : Forecast plot for regression models only available for a single predictor
```

---

```{r plot.TVreg.func, echo=FALSE}
plotforecasttv <- function(object, h=10, ylims=NULL){
  dat <- object$model
  dat$fitted <- object$fitted.values
  tlim <- (max(object$model$t)+1:h)
  pr95 <- predict(model, newdata = data.frame(t=(max(object$model$t)+1:h)), level=0.95, interval="prediction")
  pr80 <- predict(model, newdata = data.frame(t=(max(object$model$t)+1:h)), level=0.80, interval="prediction")
  pr95 <- as.data.frame(pr95); pr95$t <- tlim
  pr80 <- as.data.frame(pr80); pr80$t <- tlim
  if(is.null(ylims)) ylims <- c(min(dat[,1],pr95$lwr, pr80$lwr), max(dat[,1],pr95$upr, pr80$upr))
  p1 <- ggplot(dat, aes_string(x = colnames(dat)[2], y = colnames(dat)[1])) +
  theme_bw() +
  geom_point(color = "blue") + xlim(0,max(tlim)) + ylim(ylims) +
  geom_line(aes(x=t, y=fitted), color="red")
  
  p1 + 
    geom_ribbon(mapping=aes(x=t, ymin=lwr, ymax=upr), data=pr95, inherit.aes=FALSE, fill = "grey50") +
    geom_ribbon(mapping=aes(x=t, ymin=lwr, ymax=upr), data=pr80, inherit.aes=FALSE, fill = "grey75") +
    geom_line(aes(x=t, y=fit), pr95)
}
```

I created a function that you can use to plot time-varying regressions with polynomial $t$.  You will use this function in the lab.

```{r plot.TVreg.forecast2, fig.align = "center"}
plotforecasttv(model, ylims=c(8,17))
```

---

A feature of a time-varying regression with many polynomials is that it fits the data well, but the forecast quickly becomes uncertain due to uncertainty regarding the polynomial fit.  A simpler model can give forecasts that do not become rapidly uncertain.

The flip-side is that the simpler model may not capture the short-term trends very well and may suffer from autocorrelated residuals.



```{r tvreg.lm1}
model <- lm(log.metric.tons ~ t + I(t^2), data=sardine)
```

---

```{r plot.TVreg.lm1, fig.align = "center"}
plotforecasttv(model, ylims=c(8,17))
```

---

# Summary

* Time-varying regression is a simple approach to forecasting that allows a non-linear trend.
* The uncertainty in your forecast is determined by how much error there is between the fit an the data.
* Fit must be balanced against prediction uncertainty.
* R allows you to quickly fit models and compute the prediction intervals.

Careful thought must be given to selecting the polynomial order.

* Standard methods are available in R for order selection
* Using different orders for different data sets has prediction consequences


