---
output:
  xaringan::moon_reader:
    css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

layout: true

.hheader[<a href="index.html">`r fontawesome::fa("home", fill = "steelblue")`</a>]

---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
library(huxtable)
```

class: center, middle, inverse
# ARIMA Models
## Fitting Models

.futnote[Eli Holmes, NOAA Fisheries]

.citation[eli.holmes@noaa.gov]

---

```{r load_data, echo=FALSE, message=FALSE, warning=FALSE}
load("landings.RData")
landings$log.metric.tons = log(landings$metric.tons)
landings = subset(landings, Year <= 1989)
anchovy = subset(landings, Species=="Anchovy")$log.metric.tons
sardine = subset(landings, Species=="Sardine")$log.metric.tons

library(ggplot2)
library(gridExtra)
library(reshape2)
library(tseries)
library(forecast)
```

---

## Fitting ARIMA Models

We are now at step B of the Box-Jenkins Method.  

A. Model form selection
    1. Evaluate stationarity and seasonality
    2. Selection of the differencing level (d)
    3. Selection of the AR level (p)
    4. Selection of the MA level (q)
    
B. **Parameter estimation**

C. Model checking

---

## Fitting with `auto.arima()`

Let's try fitting to some simulated data.  We will simulate with `arima.sim()`.  We will specify no differencing.

```{r fitting.example.1}
set.seed(100)
a1 = arima.sim(n=100, model=list(ar=c(.8,.1)))
auto.arima(a1, seasonal=FALSE, max.d=0)
```

The 'best-fit' model is simpler than the model used to simulate the data. 

---

## Let's fit 100 and see how often the 'true' model is chosen

By far the correct type of model is selected, AR-p, but usually a simpler model of AR-1 is chosen over AR-2 (correct) most of the time.

```{r fit.1000}
save.fits = rep(NA,100)
for(i in 1:100){
  a1 = arima.sim(n=100, model=list(ar=c(.8,.1)))
  fit = auto.arima(a1, seasonal=FALSE, max.d=0, max.q=0)
  save.fits[i] = paste0(fit$arma[1], "-", fit$arma[2])
}
table(save.fits)
```

---

# Missing values

These fitting functions will accept missing values.

```{r fitting.example.2.data}
set.seed(100)
a1 = arima.sim(n=100, model=list(ar=c(.8,.1)))
a1[sample(100,50)]=NA
plot(a1, type="l")
title("many missing values")
```

```{r fitting.example.2}
auto.arima(a1, seasonal=FALSE, max.d=0)
```

```{r fitting.example.3}
Arima(a1, order = c(2,0,0))
```

---

## Summary

- `auto.arima()` in the forecast package is a good choice for selection and fitting of ARIMA models.

- `Arima()` is a good choice when you know the order (structure) of the model.

- You need to know whether to the mean of the data should be zero and whether it is stationary around a linear line.
    - `include.mean=TRUE` means the mean is not zero
    - `include.drift=TRUE` means fit a model that fluctuates around a line (up or down)

### Seasonality

These functions will also fit seasonal models.  We will address seasonality later.