---
output:
  xaringan::moon_reader:
    css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

layout: true

.hheader[<a href="index.html">`r fontawesome::fa("home", fill = "steelblue")`</a>]

---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
library(huxtable)
```

class: center, middle, inverse
# Forecasting Time Series
## Time-varying Regression: Forecasting

.futnote[Eli Holmes, NOAA Fisheries]

.citation[eli.holmes@noaa.gov]

---

```{r load_data_TV_Regression_Forecasting, echo=FALSE}
load("landings.RData")
landings$log.metric.tons = log(landings$metric.tons)
landings = subset(landings, Year <= 1989)
landings$t = landings$Year-landings$Year[1]
anchovy = subset(landings, Species=="Anchovy" & Year <= 1987)
sardine = subset(landings, Species=="Sardine" & Year <= 1987)
```

Forecasting is easy in R once you have a fitted model.  Use the model with the covariates $t+1$, $t+2$, etc.

For the anchovy, we fit the model

$$C_t = \alpha + \beta t + e_t$$
where $t$ is year starting at 0 (so 1963 is $t=0$).  To predict, predict the catch in year X, we use

$$C_t = \alpha + \beta X + e_t$$

---

Model fit:

```{r tvreg.anchovy2}
model <- lm(log.metric.tons ~ t, data=anchovy)
coef(model)
```

For anchovy, the estimated $\alpha$ (Intercept) is `r coef(model)[1]` and $\beta$ is `r coef(model)[2]`.  We want to use these estimates to forecast 1988 ($t=24$).

So the 1988 forecast is `r coef(model)[1]` + `r coef(model)[2]` $\times$ 24 :

```{r tvreg.forecast1}
coef(model)[1]+coef(model)[2]*24
```

log metric tons.

---

# The forecast package

The forecast package in R makes it easy to create forecasts with fitted models and to plot those forecasts.

For a TV Regression model, our `forecast()` call looks like

```{r TVregression.forecast2}
library(forecast)
fr <- forecast(model, newdata = data.frame(t=24:28))
fr
```

---

The dark grey bands are the 80% prediction intervals and the light grey are the 95% prediction intervals.

```{r plot.TVreg.forecast}
plot(fr)
```

---

Sardine forecasts from a 4th order polynomial can similarly be made:

```{r tvreg.sardine2}
model <- lm(log.metric.tons ~ t + I(t^2) + I(t^3) + I(t^4), data=sardine)
fr <- forecast(model, newdata = data.frame(t=24:28))
fr
```

---

Unfortunately, forecast does not recognize that there is only one predictor $t$
```{r plot.TVreg.forecast.bad}
try(plot(fr))
```

---

```{r plot.TVreg.func, echo=FALSE}
plotforecasttv <- function(object, h=10){
  dat <- object$model
  tlim <- (max(object$model$t)+1:h)
  pr95 <- predict(model, newdata = data.frame(t=(max(object$model$t)+1:h)), level=0.95, interval="prediction")
  pr80 <- predict(model, newdata = data.frame(t=(max(object$model$t)+1:h)), level=0.80, interval="prediction")
  pr95 <- as.data.frame(pr95); pr95$t <- tlim
  pr80 <- as.data.frame(pr80); pr80$t <- tlim
  ylims <- c(min(dat[,1],pr95$lwr, pr80$lwr), max(dat[,1],pr95$upr, pr80$upr))
  p1 <- ggplot(dat, aes_string(x = colnames(dat)[2], y = colnames(dat)[1])) +
  theme_bw() +
  geom_point(color = "blue") + xlim(0,max(tlim)) + ylim(ylims)
  p1 + 
    geom_ribbon(mapping=aes(x=t, ymin=lwr, ymax=upr), data=pr95, inherit.aes=FALSE, fill = "grey50") +
    geom_ribbon(mapping=aes(x=t, ymin=lwr, ymax=upr), data=pr80, inherit.aes=FALSE, fill = "grey75") +
    geom_line(aes(x=t, y=fit), pr95)
}
```

I created a funciton that you can use to plot time-varying regressions with polynomial $t$.  You will use this function in the lab.

```{r plot.TVreg.forecast2}
plotforecasttv(model)
```
---

