---
output:
  xaringan::moon_reader:
    css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

layout: true

.hheader[<a href="index.html">`r fontawesome::fa("home", fill = "steelblue")`</a>]

---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
knitr::opts_chunk$set(fig.height=5)
library(huxtable)
```

class: center, middle, inverse
# Developing Forecasting Models for Fisheries Time Series with R
## PSAW 2019 --- 11 Feb 2019

.futnote[Eli Holmes, NOAA Fisheries]

.citation[eli.holmes@noaa.gov]

---

## Today's objective

* Learn the Box-Jenkins method for developing a ARIMA forecasting model in R

* Learn how to develop an exponential smoothing (ETS) model for forecasting in R

* Main R package you will learn is **forecast** by Rob Hyndman, Monash University, Australia.  You will also use the base time series functions in R but we will focus on **forecast**.

* You will be working with univariate data with no covariates.

---

## Non-seasonal Data

No season. All years. If data are missing, then they are entered as NA.

```{r}
load("landings.RData")
class(landings)
head( subset(landings, Species=="Anchovy") )
```

---

## Seasonal Data

No season. All years and seasons (months, quarters, hours). If data are missing, then they are entered as NA.  Only one season.  So not daily plus monthly cycle.

```{r}
load("chinook.RData")
head( subset(chinook, Species=="Chinook") )
```

---

## State-space time series models (Mark's workshop) vs ARIMA

Let's imagine that we can describe our data as a combination of the mean trend and error.

$$x_t = m_t + e_t$$

What if our main goal is to predict $x_{t+1}$?  

Note we (fisheries biologists and ecologists) often want to know $m_t$. In fact, that is often our main and sole goal.  But let's say our goal is to predict $x_{t+1}$.

---

## Approach #1 State-space models as applied in Mark's workshop

[Time Series Analysis, PSAW II AM](https://mdscheuerell.github.io/PSAW2/)

We estimate $m_t$ from the data.  Once we have an estimate of $m_t$, we have an estimate of $e_t$.  We can model that error (think AR and MA) to help us predict future $x_t$ from $x_{t-1}$.

We will have a model for the $m_t$ and we will fit
$$x_t = m_t + e_t$$ 
directly.  We will get an estimate of the model that produced $m_t$.   The $x_t$ and $m_t$ models will have a biological interpretation.

---

So for example if our model is

$$m_t = m_{t-1}+\mu \\ x_t = m_t + e_t, \,\, e_t \sim N(0,\sigma)$$

Then we fit that model without taking any transforming our data.  

Note the above is just a linear deterministic trend observed with error, i.e. a linear regression against time.

---

## Approach #2 Box-Jenkins

The Box-Jenkins approach is different. In this approach to predicting $x_{t+1}$, we remove $m_t$ from our data using differencing.  We don't have to worry about a model for $m_t$ because we have removed it!  

And we don't actually need $m_t$ to predict $x_{t+1}$ since we have $x_{t}$, $x_{t-1}$, etc.  

We model $\Delta x_t$ and then predict $\Delta x_{t+1}$. $\Delta x_t$ is $x_t-x_{t-1}$. Once we have that we can predict $x_{t+1}$ by using $x_t$, $x_{t-1}$, etc with some algebra.

*We might also use a d-th difference.

---

The error structure of $\Delta x_{t+1}$ is NOT the same as $e_t$. 
$$\Delta x_{t} = \phi_1\Delta x_{t-1} + \phi_2\Delta x_{t-2} + \dots + z_t$$

This is a model for the DIFFERENCES and $z_t$ is the error of the differences. It is more difficult to make a biological interpretation of the ARMA model.  But remember, the objective was to predict $x_{t+1}$ not to fit a model with a biological interpretation.

---

## Examples of differencing, linear trend

$m_t$ is linear and deterministic.

$$x_t = \mu t + w_t$$

```{r, echo=FALSE}
set.seed(123)
n=100
sd=sqrt(.1)
mu=0.02
mt=mu*(1:n)
et=rnorm(n,0,sd)
xt=mt+et
plot(xt[1:100],type="l")
```

---

When we take the first difference, we get
$$\Delta x_t = \mu (t - (t-1)) + w_t - w_{t-1} = \mu + w_t - w_{t-1}$$

$w_t - w_{t-1}$ are now the errors in the model.

---

## More resources

Rob Hyndman book on forecasting and the **forecast** package: [Forecasting Principals and Practice](https://otexts.org/fpp2/)

NWFSC Time Series Analysis: https://nwfsc-timeseries.github.io/  (PIs: Eli Holmes, Mark Scheurell, Eric Ward)

* Courses on multivariate and univariate time series analysis applied to messy fisheries and ecological data.

* R packages for fitting time series models.  Bayesian and maximum-likelihood.

* online books:

  - [Applied Time-Series Analysis for Fisheries and Environmental Data](https://nwfsc-timeseries.github.io/atsa-labs/)
  
  - [Fisheries Catch Forecasting](https://fish-forecast.github.io/Fish-Forecast-Bookdown/)
