---
title: "Set-up"
---

```{r tangle_R_files, include=FALSE, echo=FALSE}
library(stringr)
# This is moving not copying for some reason
# file.copy("landings.RData",
#           "Labs/landings.RData",
#           overwrite = TRUE)
library(knitr)
library(rmarkdown)
files <- dir(pattern="^Forecasting.*Rmd$")
files <- str_sub(files,1,-5)
for(fil in files){
  infil = paste0(fil,".Rmd")
outfil = paste0("Labs/R_scripts/",fil,".R")
knit(infil, output = outfil, tangle=TRUE)
}
```

```{r render_Rmd_files, include=FALSE, echo=FALSE}
require(stringr)
files <- dir("Labs", pattern="*Rmd$",full.names=TRUE)
files <- str_sub(files,1,-5)
for(fil in files){
  infil = paste0(fil,".Rmd")
outfil = paste0(fil,".html")
render(infil, output_file=outfil)
}
```


Set-up:

1. Go to https://github.com/eeholmes/Catch-Forecasting-Files and copy all the files into a directory.

- If GitHub Desktop works for you, then you can click 'Clone or download' and then click open in Desktop.

- If GitHub Desktop is not working for you, then click 'Clone or download' and then download a zip file and unzip on your computer.

2. Open RStudio and 'New Project' in the upper right.  Then select 'From Existing Directory' and select the directory where the 'Catch-Forecasting-Files' files are.

3. Next time you open RStudio, you can select your project from the upper right..

You will need to install the following packages: 

```
install.packages("ggplot2")
install.packages("gridExtra")
install.packages("reshape2")
install.packages("urca")
install.packages("tseries")
install.packages("forecast")
```
