---
title: ARMA Models Lab 2 Extras
output:
  html_document:
    toc: true
    toc_float: true
---

This goes into some of the material in Lab 2 in more depth.

```{r load_data_Lab2_Extras}
data(greeklandings, package="FishForecast")
landings = subset(greeklandings, Year <= 1987)
anchovy = ts(subset(landings, Species=="Anchovy")$log.metric.tons, start=1964)
```

```{r load_packages}
require(urca)
require(ggplot2)
require(tseries)
require(forecast)
require(reshape2)
require(gridExtra)
```



# A random walk

A random walk is non-stationary.  The variance increases with time.  We can see this is we plot many random walks on top of each other.

A single random walk can be made with a simple for loop or with a `cumsum` on white noise.

```{r rw}
# this is a random walk
TT <- 100
sd <- 0.2
rw <- rep(0,TT)
for(i in 2:TT) rw[i] <- rw[i-1]+rnorm(1, sd=sd)
# and this is a random walk
rw <- cumsum(rnorm(TT, sd=sd))
```

Let's plot 100 random walks on top of each other.  Notice how the variance (range of values on the y-axis grows) the farther we get from the start.

```{r rws}
nsim <- 100
plot(rw, type="l", ylim=c(-4,4))
for(i in 1:nsim){
  rw <- cumsum(rnorm(TT, sd=sd))
  lines(rw)
}
```

*Optional* We can make a similar plot with ggplot.

```{r rw.ggplot}
rw <- rep(0,TT)
err <- rnorm(TT, sd)
for(i in 2:TT) rw[i]=rw[i-1]+err[i]
dat <- data.frame(t=1:TT, rw=rw)
p1 <- ggplot(dat, aes(x=t, y=rw)) + geom_line() + 
  ggtitle("Random Walk") + xlab("") + ylab("value")
rws <- apply(matrix(rnorm(TT*nsim),TT,nsim),2,cumsum)
rws <- data.frame(rws)
rws$id <- 1:TT

rws2 <- melt(rws, id.var="id")
p2 <- ggplot(rws2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("The variance of a random walk process grows in time")
grid.arrange(p1, p2, ncol = 1)
```

## *Optional* Adding an intercept to a random walk model

When we add an intercept to a random walk, it behaves very differently than a stationary process with intercept added.

$$x_t = x_{t-1} + \alpha + e_t$$

looks very different than

$$x_t = \beta_1 x_{t-1} + \alpha + e_t$$

Adding an intercept to a random walk leads to an upward linear drift.  While in the AR(1) model, it leads to a flat level of $\alpha/(1-\beta_1)$.

```{r rwt.vs.ar1t}
TT <- 100
intercept <- 1
trend <- beta1 <- 0.5
rwi <- rep(0,TT)
err <- rnorm(TT, sd=sd)
for(i in 2:TT) rwi[i] <- rwi[i-1] + intercept + err[i]
ar1i <- rep(intercept/(1-beta1),TT)
for(i in 2:TT) ar1i[i] <- beta1 * ar1i[i-1] + intercept + err[i]
par(mfrow=c(1,2))
plot(rwi, type="l")
plot(ar1i, type="l")
```


```{r rwt.vs.ar1t.ggplot}
require(ggplot2)
require(gridExtra)
dat <- data.frame(t=1:TT, y=cumsum(rnorm(TT)))
dat$yi <- cumsum(rnorm(TT,intercept,1))
dat$yti <- cumsum(rnorm(TT,intercept+trend*1:TT,1))
p1 <- ggplot(dat, aes(x=t, y=y)) + geom_line() + ggtitle("Random Walk")
p2 <- ggplot(dat, aes(x=t, y=yi)) + geom_line() + ggtitle("with non-zero mean added")
p3 <- ggplot(dat, aes(x=t, y=yti)) + geom_line() + ggtitle("with linear trend added")

grid.arrange(p1, p2, p3, ncol = 1)
```



## Dickey-Fuller test

The null hypothesis for this test is that the data are a random walk (non-stationary).  You want the null hypothesis to be rejected.

### Dickey-Fuller test on white noise

Let's first do the test on data we know is stationary, white noise.  We have to make choose the `type` and `lags`.  If you have no particular reason to not include an intercept and trend, then use `type="trend"`.  This allows both intercept and trend.  Next you need to chose the `lags`.  The default of `lags=1` is generally ok but we will use `lags=0` to fit a simpler model given that we don't have many years of data.

It is fitting this model to the data and you are testing if `z.lag.1` is 0.  

`x(t) - x(t-1) = z.lag.1 * x(t-1) + intercept + tt * t`

If you see `***` or `**` on the coefficients list for `z.lag.1`, it indicates that `z.lag.1` is significantly different than 0 and this supports the assumption of stationarity.

The `intercept` and `tt` estimates indicate where there is a non-zero level (intercept) or linear trend (tt).

```{r df.wn}
require(urca)
wn <- rnorm(TT)
test <- ur.df(wn, type="trend", lags=0)
summary(test)
```

The coefficient part of the summary indicates that `z.lag.1` is different than 0 (so stationary) an no support for intercept or trend.

Scroll down and notice that the test statistic is LESS than the critical value for `tau3` at 5 percent.  This means the null hypothesis is rejected at $\alpha=0.05$, a standard level for significance testing.


### Dickey-Fuller on the anchovy data

```{r df.anchovy}
test <- ur.df(anchovy, type="trend", lags=0)
summary(test)
```

Scroll down to the critical values.  The test-statistic is greater than the critical values.  This means that we cannot reject the null hypothesis that the data contain a random walk.


