---
title: ETS Models Lab 1
output:
  html_document:
    toc: true
    toc_float: true
---

In this lab, you will fit a series of candidate models and prepare a table for their performance.

# Set-up

We will define the training data as 1964 to 1987 and the test data as 1988 and 1989.

```{r read_data, eval=FALSE}
load("landings.RData")
spp <- "Anchovy"
landings$log.metric.tons <- log(landings$metric.tons)
landings <- subset(landings, Species==spp)
traindat <- subset(landings, Year <= 1987)$log.metric.tons
testdat <- subset(landings, Year == 1988 | Year==1989)$log.metric.tons
```

Load the necessary packages.

```{r load_forecast}
library(forecast)
library(tseries)
```

## Fit each of our candidate models

We will store these in a list for easy access.

```{r}
fit.list <- list()
fr.list <- list()
```

* Exponential smoothing model with trend

```{r}
modelname <- "ETSwtrend"
fit <- ets(traindat, model="AAN")
fit.list[[modelname]] <- fit
fr.list[[modelname]] <- forecast(fit, h=1)
```

* Exponential smoothing model no trend
```{r}
modelname <- "ETSnotrend"
fit <- ets(traindat, model="ANN")
fit.list[[modelname]] <- fit
fr.list[[modelname]] <- forecast(fit, h=1)
```

* ARIMA(0,1,1) with drift (best)
```{r}
modelname <- "ARIMA011wdrift"
fit <- Arima(traindat, order=c(0,1,1), include.drift=TRUE)
fit.list[[modelname]] <- fit
fr.list[[modelname]] <- forecast(fit, h=1)
```

* ARIMA(2,1,0) with drift (within 2 AIC of best)
```{r}
modelname <- "ARIMA210wdrift"
fit <- Arima(traindat, order=c(2,1,0), include.drift=TRUE)
fit.list[[modelname]] <- fit
fr.list[[modelname]] <- forecast(fr, h=1)
```

* Time-varying regression with linear time
```{r}
TT <- length(traindat)
#make a data.frame for lm
dat <- data.frame(log.metric.tons=traindat, t=1:TT)
modelname <- "TVreglinear"
fit <- lm(log.metric.tons ~ t, data=dat)
fit.list[[modelname]] <- fit
fr.list[[modelname]] <- forecast(fit, newdata=data.frame(t=TT+1))
```

* Naive no trend
```{r}
modelname <- "Naive"
fit <- Arima(traindat, order=c(0,1,0))
fit.list[[modelname]] <- fit
fr.list[[modelname]] <- forecast(fit, h=1)
# or simply
# fr <- rwf(traindat, h=1)
```

* Naive with trend
```{r}
modelname <- "Naivewtrend"
fit <- Arima(traindat, order=c(0,1,0), include.drift=TRUE)
fit.list[[modelname]] <- fit
fr.list[[modelname]] <- forecast(fit)
# or simply
# fr <- rwf(traindat, drift=TRUE, h=1)
```

* Average or mean
```{r}
modelname <- "Average"
fit <- Arima(traindat, order=c(0,0,0))
fit.list[[modelname]] <- fit
fr.list[[modelname]] <- forecast(fit)
```

## Models fit

Now we can use `names()` to see the models that we have fit.  If we want to add more, we use the code above as a template.

```{r}
modelnames <- names(fit.list)
modelnames
```


